---
title: "Interactive (and Static) Mapping in R"
author: "Connor Gilroy"
date: "`r Sys.Date()`"
output: 
  html_document: 
    keep_md: true
  ioslides_presentation: default
---

## Outline

- Read Excel spreadsheet into R
- Convert geohash locations to latitude & longitude
- Filter data to look at one city
- Create interactive map using Leaflet
- (Bonus: Create static map using shapefiles and ggplot2)

## Dataset

The dataset I'm using is from a gay mobile dating app, Grindr. Grindr made available a small set of aggregated, anonymized user data for a [hackathon](https://hack4equality2016-onlineonly.devpost.com/) in September 2016.

The data don't have much of a public description, but they appear to contain data for the top 100 locations in each of 15 cities. 

```{r}
library(readxl)
data_file <- "data/Data4Equality-Aug27.xlsx"
sheet_names <- excel_sheets(data_file)
sheet_names
```

----

```{r warning=FALSE, message=FALSE}
library(magrittr)
library(dplyr)
grindr_df <- 
  lapply(sheet_names, function(x) read_excel(data_file, sheet = x)) %>%
  set_names(sheet_names) %>%
  bind_rows(.id = "city")
head(grindr_df[, 1:6])
```

## Geohash to lat/lng

Geohash is a way of encoding location data. The more digits, the more precise the location. In this case, we have geohashes of length 8, with an error of ±0.019km. We can use a library to decode the geohashes into latitudes, longitudes, and errors for each.

```{r}
library(geohash)
grindr_df_locs <-
  grindr_df %>%
  bind_cols(., gh_decode(.$geohash8))
```

----

```{r}
grindr_df_locs %>% 
  select(geohash8, lat, lng, lat_error, lng_error) %>% 
  head()
```

## Montreal

Let's look at one city, Montreal. Montreal has the second-highest number of Grindr users of the 15 cities in the dataset, after Los Angeles.

```{r}
montreal_grindr_df <- 
  grindr_df_locs %>% 
  filter(city == "Montreal")
```

## Bounding box with `sp`

Create a "bounding box" for the Montreal locations.

```{r}
library(sp)
mtl_bb <- 
  montreal_grindr_df %>% 
  select(lng, lat) %>% 
  SpatialPoints(CRS("+proj=longlat")) %>%
  bbox()
mtl_bb
```

## Map with Leaflet

`leaflet` is an R interface to the JavaScript Leaflet library, which creates interactive maps. 

First, create a base map.

```{r}
library(leaflet)
m <- 
  leaflet(montreal_grindr_df) %>% 
  fitBounds(lng1 = mtl_bb[1, 1],
            lat1 = mtl_bb[2, 1],
            lng2 = mtl_bb[1, 2],
            lat2 = mtl_bb[2, 2]) %>%
  addTiles()
```

----

```{r echo=FALSE}
m
```

## Plot data

Then, add markers, sized by the number of unique users, with popups.

```{r}
library(stringr)
m_markers <- 
  m %>% 
  addCircleMarkers(radius = ~unique_users/100,
                   popup = ~str_c(as.character(unique_users), 
                                  "unique users", sep = " "), 
                   weight = 1, opacity = .5, fillOpacity = .5)
```

----

```{r echo=FALSE}
m_markers
```

## Recap: packages used

Tidyverse: `readxl`, `magrittr`, `dplyr`, `stringr`

Spatial packages: `geohash`, `sp`, `leaflet`

## Links

Github: 

Leaflet: https://rstudio.github.io/leaflet/

## Bonus: static maps

Static maps are actually more complicated, if you want to render the base map from shapefiles. 

You need to download shapefiles (city government websites often have this data), enter projection information as the `proj4string` (if that information isn't already included), and then convert that projection into a `longlat` projection.

## Read the shapefiles

```{r}
library(rgdal)
## you also need the `rgeos` package to be installed
## shape files from http://donnees.ville.montreal.qc.ca/dataset/quartiers
montreal <- readOGR(dsn = "data/montreal/quar-ref", 
                    layer = "Quar_ref")
```

## Add projection information

In order to find the proj4 string, I googled "NAD_1983_MTM_8", which was listed under *Methodologie* on the download page with no explanation. This lead me to http://spatialreference.org/ref/epsg/2950/, where I clicked on the proj4 link and copied that string.

```{r}
montreal@proj4string <- 
  CRS("+proj=tmerc +lat_0=0 +lon_0=-73.5 +k=0.9999 +x_0=304800 +y_0=0 +ellps=GRS80 +units=m +no_defs ")
```

## Convert projection

```{r}
montreal <- spTransform(montreal, CRS("+proj=longlat"))
```

## Transform into data frame

Convert the `SpatialPolygonsDataFrame` object to a regular data frame to use with `ggplot2`, using `broom::tidy()`.

```{r}
library(broom)
library(maptools)
montreal_df <- tidy(montreal, region = "NOM_QR")
```

## Create the base plot

`ggplot2` has a dedicated geom, theme, and coordinate system for maps.

```{r}
library(ggplot2)
library(ggthemes)
montreal_plot <- 
  ggplot() + 
  geom_map(data = montreal_df, map = montreal_df, 
           aes(x = long, y = lat, map_id =id), 
           fill = NA, color = "darkgray", size = .3) + 
  expand_limits(x = montreal_df$long, y = montreal_df$lat) + 
  theme_map() + 
  coord_map()
```

----

```{r echo=FALSE}
montreal_plot
```

## Label quartiers of Montreal

Create a data frame with the centroid and name of each quartier.

The quartiers with non-ASCII characters are encoded in latin1 encoding; convert those to UTF-8.

```{r}
centroids_montreal <- as_data_frame(coordinates(montreal)) 
centroids_montreal <- 
  bind_cols(centroids_montreal, 
            as_data_frame(iconv(montreal@data$NOM_QR, "latin1", "utf8")))
## or, stringr::str_conv(montreal@data$NOM_QR, "latin1")
names(centroids_montreal) <- c("long", "lat", "id")
```

## Label quartiers of Montreal

```{r}
montreal_plot_labeled <-
  montreal_plot + 
  geom_text(data = centroids_montreal, 
            aes(label = id, x = long, y = lat), size = 1.2)
```

----

```{r echo=FALSE, dpi=288, out.width="720px"}
montreal_plot_labeled
```

## Plot Grindr data on labeled map

```{r}
montreal_plot_grindr <- 
  montreal_plot_labeled + 
  geom_point(data = montreal_grindr_df, 
             aes(x = lng, y = lat, map_id = 1, color = unique_users, 
                 size = loc_pings), 
             alpha = .8) + 
  scale_color_gradient(low = "pink", high = "darkred") + 
  scale_size_continuous(guide = guide_legend(reverse = TRUE)) + 
  theme(legend.position = c(0, 1),
        legend.justification = c(0, 1), 
        legend.box = "horizontal", 
        plot.title = element_text(size = 20)) + 
  ggtitle("Grindr pings in Montreal, QC, CA")
```

----

```{r echo=FALSE, dpi=288, out.width='720px'}
montreal_plot_grindr
```
